SIMD\slash SIMT-vectorized Sparse Chemical Kinetic Jacobian\slash Thermo-chemical Source-Term Evaluation

### 2018-03-13 paper.tex
first draft -- only needs final revisions

%====================================================================
\begin{abstract} % not to exceed 200 words
A code generation platform for single-instruction, multiple-data (SIMD) and single-instruction, multiple thread (SIMT) vectorized thermo-chemical source-term and sparse\slash dense chemical kinetic Jacobian evaluation was developed and validated for a wide range of chemical kinetic models.

\todo[inline]{this}
\end{abstract}

@@

Finally, one key-component missing in this current version of \texttt{pyJac} is vectorized sparse\slash dense linear-algebra subroutines to maximize the performance of LU-factorization and matrix-vector multiplication (commonly used in implicit-integration techniques).
Third-party\slash open-source options exist for some target-languages, e.g., cuBLAS~\cite{cublas},clBLAS\slash clSPARSE~\cite{clmath} or SuperLU~\cite{superlu99}, but these do not necessarily cover all targets\slash necessary linear-algebra operations and, in the case of the CUDA\slash OpenCL libraries, are often optimized operation on one-large matrix instead of many (relatively) smaller matrices.
The extent to which these existing programs may be used needs to be assessed and missing operations should be implemented in \texttt{loo.py} to ensure easy switching between target-languages\slash vectorization types.

\section{Conclusions}
In this work, automatically generated OpenCL codes for SIMD\slash SIMT-vectorized thermo-chemical source-term and sparse\slash dense chemical kinetic Jacobian evaluation were developed and validated against for a range of chemical kinetic models~\cite{Burke:2011fh,smith_gri-mech_30,Wang:2007,Sarathy:2013jr} and multiple CPUs\slash GPUs (\cref{t:cpus,t:gpus}).
Significant speedups in shallow SIMD-vectorized source-term evaluation over a parallel OpenMP code were demonstrated, up to \SIrange{2.53}{2.92}{$\times$} and \SIrange{3.40}{4.08}{$\times$} for \sse/ and \avx/ capable CPUs; in addition, source-term evaluation was \SIrange{1.40}{1.88}{$\times$} faster on a \gpunew/ GPU compared to a \gpuold GPU; deep-vectorization, though possible on the Portable OpenCL (POCL) platform~\cite{poclIJPP}, did not yield any performance benefit as POCL did not achieve vectorizations for any execution pattern studied.
Vectorized data-orderings specifically formulated for shallow\slash deep SIMD\slash SIMT vectorizations were proposed, resulting in a shallow-vectorized ``C'' (row major) speedup of \SI{2.14}{2.58}{$\times$} over a ``F'' (column major) ordered shallow-vectorization, compared to a speedup of \SI{1.35}{2.09}{$\times$} for the ``C''-ordered OpenMP code (over an ``F''-ordered equivalent).
In addition, the effect of the \conp/\slash\conv/ formulation was determined to be negligible on source-term evaluations, while varying the vector-width size had minimal (\textasciitilde\SI{10}{$\percent$}) performance effects on both the CPU and GPU.
The strong parallel scaling efficiency was examined for source-term evaluation; the parallel OpenMP code generally had better efficiencies, but when compared to a shallow-vectorized code with roughly equivalent performance---e.g., source-term shallow-vectorized evaluation on \num{4} cores to \num{16} OpenMP cores---the scalings were similar.

>>>>

%====================================================================
\begin{abstract} % not to exceed 200 words
A code generation platform, \texttt{pyJac}, for single-instruction, multiple-data (SIMD) and single-instruction, multiple thread (SIMT) vectorized constant-pressure\slash constant-volume thermo-chemical source-term and sparse\slash dense chemical kinetic Jacobian evaluation was developed and validated for a wide range of chemical kinetic models.
Speedups of \SIrange{3.40}{4.08}{$\times$} were found for shallow-vectorized OpenCL source-rate evaluation compared a parallel OpenMP code on an \avx/ CPU, increasing to \SIrange{6.63}{9.44}{$\times$} and \SIrange{3.03}{4.23}{$\times$} for sparse and dense chemical kinetic Jacobian evaluation; in addition, the performance of a \gpunew/ GPU was compared to a \gpuold/ finding speedups of \SIrange{1.40}{1.88}{$\times$}, \SIrange{1.10}{1.59}{$\times$}, and \SIrange{1.36}{3.0}{$\times$} for source-rate, sparse Jacobian and dense Jacobian evaluation, respectively.
Additionally, the effect of data-ordering was investigated and an ordering specifically formulated for vectorized evaluation proposed; as well, the effect of the \conp/\slash\conv/ assumptions and varying vector widths were studied on source-term evaluation performance.
Speedups over a first-order finite-difference approach reached \SIrange{17.60}{45.13}{$\times$} and \SIrange{55.11}{245.63}{$\times$} for sparse\slash dense evaluation on the CPU\slash GPU respectively, while dense Jacobian evaluation was up to \SI{19.56}{$\times$} and \SI{2.84}{$\times$} times faster than a previous version of \texttt{pyJac} on a \sse/ CPU and \gpuold/ GPU respectively.
Finally, future directions for vectorized chemical kinetic evaluation and sparse linear-algebra techniques were discussed.
\end{abstract}

@@

Finally, one key-component missing in this current version of \texttt{pyJac} is vectorized sparse\slash dense linear-algebra subroutines to maximize the performance of LU-factorization and matrix-vector multiplication (commonly used in implicit-integration techniques).
Third-party\slash open-source options exist for some target-languages, e.g., cuBLAS~\cite{cublas},clBLAS\slash clSPARSE~\cite{clmath} or SuperLU~\cite{superlu99}, but these do not necessarily cover all targets\slash required linear-algebra operations and, in the case of the CUDA\slash OpenCL libraries, are often optimized operation on one-large matrix instead of many (relatively) smaller matrices.
The extent to which these existing programs may be used needs to be assessed and missing operations should be implemented in \texttt{loo.py} to ensure easy switching between target-languages\slash vectorization types.

\section{Conclusions}
In this work, automatically generated OpenCL codes for SIMD\slash SIMT-vectorized thermo-chemical source-term and sparse\slash dense chemical kinetic Jacobian evaluation were developed and validated against for a range of chemical kinetic models~\cite{Burke:2011fh,smith_gri-mech_30,Wang:2007,Sarathy:2013jr} and multiple CPUs\slash GPUs (\cref{t:cpus,t:gpus}).
Significant speedups in shallow SIMD-vectorized source-term evaluation over a parallel OpenMP code were demonstrated, up to \SIrange{2.53}{2.92}{$\times$} and \SIrange{3.40}{4.08}{$\times$} for \sse/ and \avx/ capable CPUs; in addition, source-term evaluation was \SIrange{1.40}{1.88}{$\times$} faster on a \gpunew/ GPU compared to a \gpuold/ GPU; deep-vectorization, though possible on the Portable OpenCL (POCL) platform~\cite{poclIJPP}, did not yield any performance benefit as POCL did not achieve vectorizations for any execution pattern studied.
Vectorized data-orderings specifically formulated for shallow\slash deep SIMD\slash SIMT vectorizations were proposed, resulting in a shallow-vectorized ``C'' (row major) speedup of \SIrange{2.14}{2.58}{$\times$} over a ``F'' (column major) ordered shallow-vectorization, compared to a speedup of \SIrange{1.35}{2.09}{$\times$} for the ``C''-ordered OpenMP code (over an ``F''-ordered equivalent).
In addition, the effect of the \conp/\slash\conv/ formulation was determined to be negligible on source-term evaluations, while varying the vector-width size had minimal (\textasciitilde\SI{10}{$\percent$}) performance effects on both the CPU and GPU.
The strong parallel scaling efficiency was examined for source-term evaluation; the parallel OpenMP code generally had better efficiencies, but when compared to a shallow-vectorized code with roughly equivalent performance---e.g., source-term shallow-vectorized evaluation on \num{4} cores to \num{16} OpenMP cores---the scalings were similar.
